\chapter{Analysis}
\label{chap:analysis}

In this chapter we begin by analysing two board games, namely kalah and chess to
gain a better understanding of which elements and components they contain.
First, in \secref{sec:board-game-analysis}, we define what a board game actually
is. After this analysis we, in \secref{sec:chessandkalah}, investigate theories
behind programming languages. 

After the analysis of board games, we begin by analysing programming paradigms
in \secref{sec:paradigms} and how the different paradigms are used and what they
are good for. 

After this we take a look at how syntactic analysis is done in
\secref{sec:syntacticanalysis} and which methods we can use to analyse our own
syntax. In this section we for instance look at context-free grammars in
\secref{sec:context-freegrammars} and different approaches to parsing in
\secref{subsec:llparsersandlrparsers} and how a parser can be constructed in
\secref{subsec:constructingaparser}. After this we look at which contextual
constraints context-free grammars have in \secref{sec:contextualconstraints}.

Furthermore, we take a look at code generation and interpretation of a given
source code in \secref{sec:codegenerationandinterpretation}. In
\secref{sec:translationprocess} we investigate the translation process. We also
describe when an intermediate language can be used and what it is good for in
\secref{sec:intermediatelanguage}. Lastly in this section, we define what a
compiler is in \secref{sec:compilation} and what an interpreter is in
\secref{sec:interpretation}.

In \secref{sec:simulator} we investigate and describe how a simulator can be use
to visualise the produced game in our programming language. Here we also
describe which possible features we can add to make the simulator a good
environment for the user.

This chapter must make it possible for us to make informed design decisions for
our programming language. Lastly in \secref{sec:summaryofdecisions}, we present
our major decisions which have been made throughout this chapter.

\input{content/analysis/broadanalysisofboardgames}
\input{content/analysis/chessandkalah}
\input{content/analysis/paradigms}
\input{content/analysis/syntacticanalysis}
\input{content/analysis/contextualconstraints}
\input{content/analysis/codegeneration-interpretation}
\input{content/analysis/simulator}
\input{content/analysis/summaryofdecisions}
