\chapter{Conclusion}
\label{chap:conclusion}

In this section we conclude on the curriculum of this semester and we also
verify or falsify the requirements, which we described in
\chapref{chap:requirements}. These where the requirements our programming langauge
should respect.

We have accomplished the goal of implementing a programming language in which it
is possible to program board games. The programming language is called
\productname{}.

We have respected the goals of the curriculum throughout the report. In the analysis 
chapter (\chapref{chap:analysis}) we have for instance accounted for different 
techniques behind the construction of compilation and interpretation.

In the design chapter (\chapref{chap:design}) and the implementation chapter
(\chapref{chap:implemention}) we have documented how we have designed, defined, and
implemented \productname{}. In the analysis chapter we accounted for the
different phases of translation of code, and in the design chapter we designed
and defined the programming language. After we designed \productname{}, we
documented our implementation of the langauge in the implementation chapter.

At a relatively early stage we concluded that we needed a simulator in which the
games written in \productname{} could be visualised and played. This decision
was based upon the conclusion that it would be very boring to play a game in the
console window. We constructed a game simulator which is presented in 
\secref{sec:simulator}.

We have not accomplished all the requiremnets which we set for this project:

We did not accomplish \textbf{requirement 1.a}. This was due to special rules of
Chess: the ``En Passant'' and ``Castling'' moves, which proved to be more
difficult to implement than we expected. Although, we still believe that
\productname{} provides the features necessary to implement them.

\textbf{Requirement 2-5, 7-17, 20-24} have all been accomplished.

\textbf{Requirement 6} has been accomplished. \textbf{Requirement 6.j} is not 
directly implemented but it is possible. This is explained earlier in 
\secref{sec:requirementsevaluation}.

The requirements presented in \textbf{requirement 18} were considered as
``could-haves'' and they therefore had a low priority.

Based on the games we have programmed in \productname{} (Noughts and Crosses,
and Connect Four), which were possible to create in around 20 lines of code, we
conclude that we have accomplished \textbf{requirement 19}.

The speed in which a game can be created in \productname{} depends on a number of
parameters. First of all, the programmers familiarity with the programming
language and the programmer's experience with programming overall. Also the
programming language's support for relevant functionality and its overall
writability is an important parameter. The fact that we were able to create a fully
functional Connect Four game in 5 minutes can be considered as a quick
implementation and witnesses that it is easy to develop board games in \productname{}.

\textbf{Requirement 29} has been accomplished, because all components of
\productname{} are compiled to Java byte code, which will be supported by all
machines with the Java virtual machine.

\section{Discussion}
\label{sec:discussion}

In this section we will discuss what we could have done differently throughout
the project. For instance, we will take a look at what we could have done
instead of developing an interpreter for \productname{}.

\input{content/conclusion/compiletointermediate}
\input{content/conclusion/bigintegers}
\input{content/conclusion/artefacts}
\input{content/conclusion/genericactions}
\input{content/conclusion/boardtypes}
\input{content/conclusion/patternmatchingefficiency}

