\chapter{Introduction}
\label{chap:introduction}

\section{A new programming language?}
So, are we really going to create a new programming language? At this very moment, hundreds, maybe thousands of new programming languages are beeing created. A Wikipedia article about the programming languages that exists, currently contains a list of 664 different programming language \cite{listofprogramminglanguages}, and alot more languages exists, since this is only a list of the more well-known languages. So why is yet another programming language needed when their exists so many programming languages already, and when it's possible in, programming languages like Java, C\# and others, to basicly program anything one would like (refered to as general purpose programming languages). The answer lies within the powers of domain-specific programming languages - a language designed to express solutions to problems in a specific domain \cite{domainspecificprogramminglanguagedefinition}. 

What we intend to do in this project is develop a programming language which can describe board games. That is designing it, defining it and implementing it. We choose to develop a programming language for programming board games in, because of our personal interest in board games. Board games are first of all fun and entertaining. They can also be rich on learning opportunities \cite{whyboardgames1?} and they can give a certain satisfaction \cite{whyboardgames2?}. In fact, board games can be almost anything as long as the creator of the game has the right amount of creativity and the building blocks needed to create the game. These are some of the reasons why it would be pratical with a programming language which made it easy to program board games in. Yes, it is possible to program board games in Java, C\# and the other general purpose programming languages, but is it easy? Definetly not as easy as in a well crafted programming language specifically designed to program board games in. If a language purely concentrates on allowing the programmers using it to express how his game works, without having to reimplement everything from scratch as he would in a general purpose programming language, it would be quicker and easier to develop shorter and more precise programs.  

Furthermore data structures and special statements specifically designed to help define board games would greatly increase the readability and writability of such a program. A language designed with board games in mind would allow the programmers to, relatively quickly, explore new ideas for a board game with a simple implementation. He/she could then efficiently modify the code according to a new rule or idea that suddenly came up, without having to make the whole thing over. If the language also took multiple platforms into account, it would open up the possibility to run the same game across multiple devices. So there are definitly many advantages of creating a new domain-specific programming language. 

Now the question is, how to develop a new domain-specific programming language in which one can program board games? First we will have to have to do some research about board games and the elements and components they consist of. We will have to know about the different programming paradigms there exists, we will have to know about compilers and/or interpreters and many other things. All of these things are included in \chapref{chap:analysis}

%\input{content/introduction/projectgoals}
%\input{content/introduction/whyboardgames}
