\subsection{Lexical analysis}
\label{sec:lexicalanalysis}
Before the syntax analysis can be performed a scanner must perform lexical
analysis. The scanner's job is basically to check a given source code for
lexical errors and translating the input stream of characters from the source
code into a stream of tokens which the parser can work with. This is done by
identifying every lexeme in the source and attaching a potential token to
it\cite[p. 57]{fischer2009}.

Lexemes are strings of characters described by regular expressions.  Typical
examples of lexemes in a programming language are: variable names, integer
literals, operators and special keywords etc. A variable name lexeme could be
defined by the following regular expression: $[a-z, A-Z, ``\_''][a-z, A-Z, 0-9,
``\_'']^*$. Which means a variable name can start with either an upper case
letter, lower case letter or an underscore followed by zero or more lower case
letters, upper case letters, numbers or underscores. The syntax or semantics of
regular expressions are not described here, we refer to Perl
notation\cite{perlRegex}. In \tableref{table:lexandtokens} we give examples of
lexemes and the tokens they have been paired with. If both $a$ and
$b$ are lexemes describing variable names and $102$ and
$42$ are lexemes describing integer literals, then $a$ and
$b$ or $102$ and $42$ can typically be used interchangeably
and still give a syntactic meaningful program.

\tab[10cm]{lexandtokens}{1}{Lexemes and their corresponding token group.}
		             {               }
       {Lexemes             }{\textbf{Tokens} } {
\tabrow{ x                  }{VAR\_NAME       } 
\tabrow{ random\_var\_name  }{VAR\_NAME       }
\tabrow{ RANdom\_var\_name2 }{VAR\_NAME       }
\tabrow{ 1 		    }{INT\_LITERAL    }
\tabrow{ 342 		    }{INT\_LITERAL    }
\tabrow{ 52890 		    }{INT\_LITERAL    }
\tabrow{ +		    }{PLUS\_OPERATOR  }
\tabrow{ - 		    }{MINUS\_OPERATOR }
\tabrow{ * 		    }{MULT\_OPERATOR  }
\tabrow{ if		    }{KEYWORD         }
\tabrow{ while 		    }{KEYWORD  	      }
\tabrow{ switch 	    }{KEYWORD  	      }
}

A scanner is a relatively simple component which can be constructed by writing
it by hand or using a scanner-generating tool such as Lex, which generates an
executable scanner by feeding it with a set of regular expressions. When
implementing the scanner for \productname{}, we would likely benefit more from
crafting the scanner by hand than by using a scanner-generating tool. By
crafting it by hand we will know exactly how it is implemented. There might be
some advantages of using a scanner-generating tool such as the fact that it can
run faster due to optimized code, is more reliable, it is easier to maintain and
it is faster to implement if one already knows how it works, if not, a
handwritten scanner could be just as fast to write.
