\section{Parser}
\label{sec:parserimplementation}

In this section we present \productname{}'s handwritten parser. We have
written a top-down recursive descent parser, which is within the class
of LL($1$) parsers. The grammar for \productname{} is suited for this
because it does not have left-recursive productions. In the end of
the section we present our work with SableCC and the reason why we choose
not to use this tool to make our scanner and parser.

\subsection{Constructing the parser}
%structured as the grammar
The parser was very simple to implement, because it is structured exactly the same way as the
grammar which can be found in \chapref{ap:CFG}.  For instance if the grammar
expresses that the next set of terminals must begin with a left bracket (`['),
  then the parser will expect the next token to be a \tokenref{LBRACKET} which
  is the token name for a left bracket. If the grammar then expects a
  non-terminal, then the parser simply calls the method for that non-terminal,
  which may in terms have the effect of calling more non-terminals and expecting terminals (which are recognised as tokens) at some points, 
  before continuing parsing the next part of the rule.

%discuss the if expression
In \lstref{lst:ifexpr}, we give an example of how this structure looks like in
our handwritten parser. The production rule for an if expression is presented in
\secref{sec:conditionalexpressions}.

%\begin{ebnf}
%\grule{if\_expr}{\gter{if} \gcat expression \gcat \gter{then} \gcat expression
%\gcat \gter{else} \gcat expression}
%\end{ebnf}

The production for if expression says that every expression of this type
must start with the combination of the two symbols spelling the word
\gter{if}. When the parser meets this word in an expression, it knows
that it has to parse an if expression.

\lstinputlisting[caption={How if expressions are parsed using top-down parsing in Java.},
label=lst:ifexpr, language=Java]{listings/ifexpr.java}

\subsection{Building an abstract syntax tree}
%astNode()
In \lstref{lst:ifexpr}, the parser initialises the node for the
expected if expression. The parser starts by calling the method
\methodref{astNode} to create a node for the Abstract Syntax Tree (AST).
We call the method with information about what type of expression this
is (\tokenref{IF\_EXPR}). The method calls the \methodref{expect} method
to verify that the next token is what we are expecting. If the two
tokens do not match, the parser throws a syntax error with information
about the error. If everything is syntactically correct the parser
constructs a node for the AST for the given expression. The first child
of the node is the boolean expression, and the next two siblings of that
child are the expression branches of the if expression.

\subsubsection{Terminal and nonterminals}
Every grammar has a finite set of nonterminals and terminals. 
A production rule specifies a nonterminal as a sequence of nonterminals and terminals. 
We have defined tokens in the
parser for every nonterminal in our grammar. The if expression has the
token name \tokenref{IF\_EXPR}.

In the production for the if expression, we have three terminals: the \gter{if},
\gter{then}, and the \gter{else}. These are all required in the method for any
if expression. When the parser finishes reading a terminal, it knows that the
following token will be an expression, and therefore a new child for the node is
made with a call to the \methodref{expression} method wherein we parse
expressions. Finally the method returns the node containing every child for the
whole if expression.

\subsection{Looking ahead in the input}
%lookAhead methods - atomic
We mentioned earlier that the parser is an LL($1$) parser, which means that the
parser only needs to look at the first token in the input stream before deciding what production rule to apply. 
We have shown the \methodref{lookAheadAtomic} method which determines if the next token is part of an
atomic expression. The production for the atomic expression is presented in
\secref{sec:atomicexpressions}.

%\begin{ebnf}
%\grule{atomic}{\gter{(} \gcat expression \gcat \gter{)}}
%\galt{variable}
%\galt{list}
%\galt{\gter{/} \gcat pattern \gcat \gter{/}}
%\galt{\gter{this}}
%\galt{\gter{super}}
%\galt{direction}
%\galt{coordinate}
%\galt{integer}
%\galt{string}
%\galt{type}
%\galt{constant}
%\end{ebnf}

An atomic expression can derive quite a few productions. This is why we have
constructed a specific method to determine whether the next token is part of an
atomic expression. This method is shown in \lstref{lst:lookaheadatomic}.

\lstinputlisting[caption={The lookAhead method to determine if the next
expression is an atomic type.}, label=lst:lookaheadatomic,
language=Java]{listings/method_lookAheadAtomic.java}

The method \methodref{lookAheadAtomic} makes use of two methods to figure out if
the next token is part of an atomic expression. The first method is the
\methodref{lookAhead} method that takes a token as argument and figures out if it equals 
the next token in the sequence of tokens. The second
method is the \methodref{lookAheadLiteral} method which is similar to the method 
in \lstref{lst:lookaheadatomic} but instead of checking for atomic expressions it 
checks for literals. All these methods return true or false.

%example of lookAheadAtomic
%LL(1)
In \lstref{lst:examplelookahead} we show an example of how the
\methodref{lookAhead} method is used in the parser. The example is taken from
the \methodref{expression} method. The productions for expressions are presented
in \secref{sec:expressions}.
The production of an expression is reflected in the code of the parser. An
example of this is given in \lstref{lst:examplelookahead}.

%\begin{ebnf}
%\grule{expression}{assignment}
%\galt{if\_expr}
%\galt{lambda\_expr}
%\galt{\gter{not} \gcat expression}
%\galt{operation}
%\end{ebnf}

\lstinputlisting[caption={Use of the \methodref{lookAhead} method. This example
is from the \methodref{expression} method.}, label=lst:examplelookahead,
language=Java]{listings/example_lookAheadAtomic.java}

The code presented in \lstref{lst:examplelookahead} is a small section of the
\methodref{expression} method. We have removed code from the section which is
not relevant for the example we are trying to give. The removed code is
presented as \{\ldots\}. An example is the code left out in the bottom which is 
just related to error handling.

An assignment begins with the reserved word \gter{let} and the first
\methodref{lookAhead} method peeks for that particular token to determine
if the next production is an assignment. If the method returns true, then
the next token is in fact the \gter{let} word, and the parser enters a
new method, namely the \methodref{assignment} method which parses the assignment and checks that the 
next tokens from the input complies with the production rule of an assignemnt. The same
is done for the if expression, lambda expression and operations which
begins with the ``loSequence()'' (logical operators).

The operation production is a bit different, because it needs two lookAhead
methods to determine if the next production is an operation (notice that this does not mean that it looks two tokens ahead). An operation can
begin with either an atomic value or a minus operator. So the code uses a
\methodref{lookAheadAtomic} and a regular \methodref{lookAhead} with the
specific token as a parameter to check if the next production is an operation.

The methods return nodes that may have children nodes. These nodes are all connected to form a tree with a single root. 
This tree is called an AST and consists of nodes for all nonterminals and leaves for all terminals, which is the tokens generated by the scanner. 
When the AST has been constructed, it can be examined to verify that the input program has been parsed correct. The class 
\classref{PrettyPrinter} has been implemented to output the structure and contents of an AST, making it easier to verify that the parser parses correct.

\subsection{SableCC}
We have also implemented a scanner and parser using a
compiler/interpreter generator, also known as a compiler compiler, namely
SableCC\cite{sableccdoc}. As described in
\secref{subsec:generatedparsers}, it is an automated scanner and
LALR($1$) parser generator written in Java, with support for making
compilers and interpreters. We have implemented an early version of
\productname{} in SableCC to evaluate the capabilities of such a tool.

\subsubsection{Choice of SableCC}
We chose SableCC instead of various other popular tools such as
ANTLR\cite{antlr} and JavaCC\cite{javacc}. Though ANTLR and JavaCC are
far more well-documented than SableCC, we still choose to work with
this tool because its parser generates a LALR($1$) parser which is generally a 
stronger parser than the LL(1) parser we wrote by hand or the LL($k$) parsers 
generated by ANTLR and JavaCC. One of the major advantages is the LALR-parser's 
ability to handle left-recursion, which obviates left-factorising.

SableCC outputs an abstract syntax node type for each alternation
in every rule in the grammar specifications file, which is merely a
simple text file with EBNF-like syntax. It's then possible to iterate
over these nodes by extending the visitor pattern that SableCC also
supplies, generating code or directly interpreting a syntactically
correct program. This is all done in classes separate from the grammar
specifications, which is also desirable and different from ANTLR
and JavaCC, where action code is injected directly in the grammar
specification. This is all done in Java, which is also desirable, since
it would work well with the rest of the project, written in Java.

\subsubsection{Experience with SableCC}
Our experience with the tool has been rather cumbersome, in that it took
quite a while to read the documentation before and while writing the
specifications, as it simply isn't just copy/pasting EBNF grammar
into a file. An example of the if-expression rule is seen below:

\begin{lstlisting}[caption={Part of the grammar specifications file of SableCC, with focus on if expressions.}]
Tokens
  else          = 'else';
  then          = 'then';
  if            = 'if';

Productions
  expression   = {elopexp} element operator expression
               | {assign} assignment
               | {if} if_expr
               | {lambda} lambda_expr
               | {el} element list?
               | {not} not expression;
  if_expr      = if [left]:expression then [mid]:expression else [right]:expression;
\end{lstlisting}

% Pros
This example brings out the strengths of SableCC, as it looks very
similar to the EBNF for expressions and if expressions seen in
\secref{sec:expressions}, with a few additions. As long as you know
the special syntax and how helpers, tokens, and productions works, it
is possible to create scanners and parsers very quickly. This was not
our case, as no one had experience in using a compiler-compiler. 
The fact that there's a clear and clean separation between automated, generated code
and user code makes the grammar and compilation/interpretation parts
easier to maintain. When adding new features to the language, you simply
have to update the specifications file and generate a new scanner/parser
combination. Whereas when adding new features to the language
while using a handwritten scanner and parser many lines of code needs
to be change in order to implement the new feature.

% Cons
Even though SableCC looks like a prime candidate to continue
interpretation with, we chose not to use the tool. This is because it
took an unreasonable amount of time to figure out how to precisely
define the grammar correctly in SableCC due to their poor
documentation. Also, it offers less control and
customisability, compared to writing our own from scratch. As an
example, the tool offers an application-specific interface to tree
walking the AST nodes with the visitor pattern, requiring knowledge of
how SableCC implements it. SableCC also generates around \num{17000} lines
of Java code, even for our simple grammar, which seems superfluous
compared to our handwritten scanner and parser, consisting of ``only''
around \num{1500} lines of code.
%  - Don't learn as much about different parser techniques
%  - Old project, not as active anymore

\subsubsection{Discussion of using a compiler compiler}
We chose not to continue using SableCC on our updated grammar, due
to the weight of cons against pros, and the fact that the time spent
working on implementing SableCC was also spent making the handwritten
scanner and parser and making them work exactly the way we want them to.
It might not be as quick to make changes to our language with this solution, but
the time spent on modifying and adding features to our language with the
use of a handwritten scanner and parser is not wasted time, as it was
spent learning, which gives us better understanding of their underlying
functionality.
