\section{Lexical structure}
\label{sec:lexicalstructure}

This section presents the low-level non-terminals of \productname{}.
We begin by describing the conventions we use throughout the report. Then we
describe the contents of the lexical structure of \productname{}, such as the
different reserved keywrods, identifiers, and literals that \productname{} contains.

\subsection{Notational Conventions}
We use a variant of Extended Backus-Naur Form (EBNF) to express the context-free
grammar of our programming language.

Each production rule assigns an expression of terminals, non-terminals and operations
to a non-terminal. E.g. in the following example the non-terminal $decimal$ is assigned
the possible terminals of $\gter{0}$ up to, and including, $\gter{9}$.

\begin{ebnf}
\grule{decimal}{\gter{0} \gor \gter{1} \gor \grange \gor \gter{9}}
\end{ebnf}

The following operations are used throughout this section to describe the
grammar of the programming language:

\begin{center}
\begin{tabular}{r l}
  $\gopt{pattern}$ & an optional pattern \\
  $\grep{pattern}$ & zero or more repititions of pattern \\
  $\ggrp{pattern}$ & a group \\
  $pattern_1 \gor pattern_2$ & a selection \\
  $\gter{0} \gor \grange \gor \gter{9}$ & a range of terminals \\
  $pattern_1 \gex pattern_2$ & matched by $pattern_1$ but not by $pattern_2$\\
  $pattern_1 \gcat pattern_2$ & concatenation of $pattern_1$ and $pattern_2$ \\
  $\gter{test}$ & a terminal \\
  $\gtsq$ & a terminal single quotation mark \\
  $\gtdq$ & a terminal double quotation mark \\
  $\gtbs$ & a terminal backslash character \\
\end{tabular}
\end{center}

The left-hand side is what will be seen in the grammar, while the right-hand
side is a description of what the operation means. For instance  a terminal will
always consist of $\gter{ }$ around the name of the terminal.

\subsection{Character Classes}
To be able to describe which characters or symbols a non-terminal can consist of
in a concise manner, we need to define some sets of symbols to specific names
which we can use in the description of our grammar.

The following classes of characters will be used throughout this section:

\begin{ebnf}
\grule{decimal}{\gter{0} \gor \gter{1} \gor \grange \gor \gter{9}}
\grule{lowercase}{\gter{a} \gor \gter{b} \gor \grange \gor \gter{z}}
\grule{uppercase}{\gter{A} \gor \gter{B} \gor \grange \gor \gter{Z}}
\grule{anycase}{lowercase \gor uppercase}
\grule{alphanum}{anycase \gor decimal}
\grule{quotebs}{\gtdq \gor \gtbs}
\grule{unichar}{\gcomment{any unicode character}}
\grule{strchar}{unichar \gex quotebs}
\end{ebnf}

For instance a string character $\left(strchar\right)$ can by any
unicode character besides a quotation mark or a backslash. 

\subsection{Comments}

In \productname{} a single-line comment begins with a sequence of at least two
forward slashes (\texttt{//}). The comment ends at the next newline. Everything
after the first two forward slashes and until the first next newline is
completely ignored at the leve of lexical analysis. Comments are valid
whitespace.

The following example shows a valid comment within an expression:

\codesample{validcomment.junta}

Unlike other programming languages, \productname{} does not have support for
multi-line comments (such as \texttt{/* */} in C-like languages).

\subsection{Reserved keywords}

The following list presents the keywords which are reserved in
\productname{}. These keywords cannot be used as identifiers in the language,
which is apparent in the following section.

\begin{ebnf}
  \grule{reserved}{\gter{define} \gor \gter{type} \gor \gter{abstract} \gor
        \gter{data} \gor \gter{extends} \gor \gter{let} \gor \gter{in}}
  \galt{\gter{set} \gor \gter{if} \gor \gter{then} \gor \gter{else} \gor
        \gter{not} \gor \gter{and} \gor \gter{or} \gor \gter{this} \gor \gter{super}}
  \galt{\gter{foe} \gor \gter{friend} \gor \gter{empty} \gor \gter{is}}
\end{ebnf}

\subsection{Identifiers}

\productname{} has three different identifiers which are defined as follows:

\begin{ebnf}
%Identifiers
\grule{constant}{\ggrp{lowercase \gcat \grep{alphanum}} \gex \ggrp{reserved \gor direction}}
\grule{type}{\ggrp{uppercase \gcat \grep{alphanum}} \gex coordinate}
\grule{variable}{\gter{\$} \gcat alphanum \gcat \grep{alphanum}}
\end{ebnf}

Constants, types, and variables cannot be defined with equal names because
they begin with different characters. 

\subsection{Literals}

The following list presents the literals of \productname{}:

\begin{ebnf}
%Literals
\grule{integer}{decimal \gcat \grep{decimal}}
\grule{direction}{\gter{n} \gor \gter{s} \gor \gter{e} \gor \gter{w} \gor
\gter{ne} \gor \gter{nw} \gor \gter{se} \gor \gter{sw}}
\grule{coordinate}{uppercase \gcat \grep{uppercase} \gcat decimal \gcat \grep{decimal}}
\end{ebnf}

The string literal can contain any unicode character if it is escaped. This lets
us construct any string. For instance if one needs to use a quotation mark
within a string, it is possible by escaping the quotation marks.

\begin{ebnf}
\grule{string}{\gtdq \gcat \grep{strchar \gor escape} \gcat \gtdq}
\grule{escape}{\gtbs \gcat unichar}
\end{ebnf}


