\section{Overview of the four main programming paradigms}
\label{sec:paradigms}

A programming paradigm describes a method and style of computer programming.
Some of the primary paradigms are imperative, object-oriented, functional and
declarative programming. While some programming languages strictly follow one
paradigm, there are many so-called multi-paradigm languages, that implement
several paradigms and therefore allow multiple styles of programming. Examples
of multi-paradigm languages include C\# and Java. It is essential for us to be
aware of these different paradigms, since it may help us to find a good style of
programming for the design of \productname{}.

\subsection{Overview}
The four main paradigms are described as follows:\cite{fourparadigms}
\begin{description}
\item[Imperative programming] describes computation in terms of statements that
  change the program state. Primary characteristics are assignments, procedures,
  data structures, control structures. Imperative programming can be seen as a
  direct abstraction of how most computers work, and many imperative languages
  are just abstractions of assembly language. Typical examples of imperative
  languages are C and Fortran.
\item[Object-oriented programming] describes computation in terms of objects
  described by attributes manipulated through methods. Primary characteristics
  are objects, classes, methods, encapsulation, polymorphism, inheritance. An
  example of a pure object-oriented languages is Smalltalk, while many other
  languages are either primarily designed for object-oriented programming (such
  as Java and C\#) or have support for object-oriented programming (such as PHP
  and Perl).
\item[Declarative programming] describes computational logic without describing
  control flow, i.e. describing {\em what} a program does rather than {\em how}
  it does it. Many domain-specific languages such as SQL, HTML and CSS are
  declarative. Logic programming, such as Prolog, is a subset of declarative
  programming.
\item[Functional programming] describes computation in terms of mathematical
  functions and seeks to avoid program state and mutable data. Purely functional
  functions have no side effects, and the result is constant in relation to the
  parameters (e.g. $add(2, 4)$ always returns $6$). An example of a purely
  function programming languages is Haskell. Other examples of languages
  designed for functional programming are Erlang, F\# and Lisp, while it is
  possible to apply functional programming concepts to many other languages.
\end{description}


While general-purpose languages, such as C\# and Java, generally tend to lean
towards the imperative and object-oriented paradigms, a domain-specific
language, with very specific goals in design, may benefit from other paradigms,
e.g. declarative programming. We want to be able to have complete control
over state changes in board games, so following the functional paradigm
and preventing side-effects, will make it easier to manage state changes
and the movement history of a game. Also, the object-oriented paradigm,
has proved to be very useful for describing many real-word systems. Allowing
the use of classes and objects, may make it easier to create board games
(e.g. an instance of class \emph{Board} consists of many \emph{Squares} 
containing \emph{Pieces} belonging to \emph{Players}). Combining these two
paradigms, we will get a functional object-oriented programming language
for describing board games.

